### 2017-05-04 2017mergingmerges/section/03-Definition.tex
serious cutting

A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the instruction currently being executed, and the @instrs@ a map of labels to instructions. We use the same record when specifying both the definition of a particular process, as well as when giving the evaluation semantics. When specifying a process the @label@ field gives the entry-point to the process code, though during evaluation it is the label of the instruction currently being executed. Likewise, when specifying a process we usually only list channel names in the @ins@ field, though during evaluation they are also paired with their current $\InputState$. If an $\InputState$ is not specified we assume it is `none'.

@@

All instructions include a $\Next$ field which is a pair of the label of the next instruction to execute, as well as a list of $\Var \times \Exp$ bindings used to update the heap. The list of update bindings is attached directly to instructions to make the fusion algorithm easier to specify, in contrast to a presentation with a separate @update@ instruction.

@@

\begin{enumerate}
\item \emph{Injection} of a single value from a stream into a process, or a set of processes. Each individual process only needs to accept the value when it is ready for it, and injection into a set of processes succeeds only when they \emph{all} accept it.

\item \emph{Advancing} a single process from one state to another. Advancing a set of processes suceeds when \emph{any} of the processes in the set can advance.

@@

Execution of a process network is non-deterministic. At any moment several processes may be able to take a step, while others are blocked. As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking. This enables the overall sequence of values on each output channel to be deterministic. Unlike Kahn processes, pushing to a channel can also block. Each consumer has a single element buffer, and pushing only succeeds when that buffer is empty.

@@

Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions is not. When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the process network to advance without requiring unbounded buffering.

@@

% -----------------------------------------------------------------------------
\subsubsection{Injection}
Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes. The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''. The @injects@ form is similar, operating on sets of processes.

@@

Rule (InjectMany) attempts to inject a single value into a set of processes. We use the single process judgment form to inject the value into all processes in the set, which must succeed for all of them. Once a value has been injected into all consuming processes that require it, the producing process no longer needs to retain it.

@@

Rule (Pull) takes the @pending@ value $v$ from the channel state and produces a heap update that will copy this value into the variable $x$ named in the @pull@ instruction. We use the syntax $us,x=v$ to mean that the list of updates $us$ is extended with the new binding $x=v$. In the result channel states, the state of the channel $c$ that was pulled from is set to @have@, to indicate the value has been copied into the local variable.

@@

The statement $\ProcShake{p}{a}{p'}$ reads ``process $p$ advances to new process $p'$, yielding action $a$''. Rule (Advance) advances a single process. We lookup the current instruction pointed to by the processes @label@ and pass it, along with the current channel states and heap to the previous single instruction judgment. The update expressions @us@ that the single instruction judgment yields are first reduced to values before updating the heap. We use $(us \lhd bs)$ to replace bindings in $us$ with new ones from $bs$. The update expressions themselves are all pure, so the evaluation can safely be done in parallel (or in arbitrary order).

% -----------------------------------------------------------------------------
\subsubsection{Feeding}
Fig.~\ref{fig:Process:Eval:Feed} gives the rules for collecting output actions and feeding the contained values to other processes. The first set of rules concerns feeding values to other processes within the same process network, while the second exchanges input and output values with the environment the process network is running in.

The statement $\ProcShake{ps}{a}{ps'}$ reads ``the processes group $ps$ advances to the new process group $ps'$ yielding output action $a$. A process ``group'' is just a set of processes.

%%% AR: why `group' and not `network'? I think network, as in Kahn process network, is fairly standard terminology.
%%% BL: Structurally it's just a set of processes, but I at some point I used "set" for setting bindings in the heap. If we change group -> network make sure to get all occurrences in the paper.

Rule (ProcessInternal) allows an arbitrary process in the group to advance to a new state at any time, provided it does not yield an output action. This allows processes to perform internal computation, without needing to synchronize with the rest of the group.

Rule (ProcessPush) allows an arbitrary process in the group to advance to a new state while yielding an output action (@push@ c v). For this to happen it must be possible to inject the output value @v@ into all processes that have channel @c@ as one of their inputs. As all consuming processes must accept the output value at the time it is created, there is no need to buffer it further in the producing process. When any process in the group produces an output action we take that as the action of the whole group.

\smallskip
The statement $\ProcsFeed{cvs}{ps}{cvs'}{ps'}$ reads ``with channel values $cvs$, process group $ps$ takes a step and produces new channel values $cvs'$ and group $ps'$''. The channel values $cvs$ map channel names to a list of values. For input channels of the overall group, we initialize the map to contain a list of input values for each channel. For output channels of the overall group, values pushed to those channels are also collected in the same channel map. In a concrete implementation the input and output values would be transported over some IO device, but for the semantics here is sufficient to describe the abstract behavior of our system.

Rule (FeedInternal) allows the process group to perform local computation in the context of the channel values.

Rule (FeedPush) collects an output action (@push@ $c$ $v$) produced by a process group and appends the value $v$ to the list corresponding to the output channel $c$.

Rule (FeedExternal) injects values from the external environment. This rule also has the side condition that values cannot be injected from the environment into output channels that are already owned by some process. This constraint is used during formal proofs of correctness, but would not need to be checked dynamically in a concrete implementation. The topology of the dataflow network does not change at runtime, so it only needs to be checked once, before execution.

% -----------------------------------------------------------------------------
\subsection{Non-deterministic Execution Order}
\label{s:EvaluationOrder}
%%% AR: Can we simplify this?
% Maybe we can start by talking about the different evaluation orders for just two alt2s:
%
% let s1 = alt2 sInA sInB
% let s2 = alt2 sInA sInC
%
% Here we have many evaluation orders. If we just look at the order in which inputs sInA etc are pulled, some options are:
% [ A, A, B, C, B, C ]
% [ A, A, B, B, C, C ]
% [ A, A, C, C, B, B ]
%
% Now if we add a new operator which zips the two together, it actually restricts the pulling order
%
% let s1 = alt2 sInA sInB
% let s2 = alt2 sInA sInC
% let s3 = zip  s1 s2
%
% This restricts the pulling order to:
% [ A, A, B, C, B, C ]
%
%%%
% Or, maybe we don't even need the zip here, and can leave the zip until S5.1.
%%%

The execution rules of Fig. \ref{fig:Process:Eval:Feed} are non-deterministic in several ways. Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the group. More importantly, (ProcessPush) allows any process to perform a push action at any time, provided all other processes in the group are ready to accept the pushed value. Rule (FeedExternal) also allows new values to be injected from the environment, provided all processes that use the associated channel are ready to accept the value.

In our system, allowing the execution order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering. Consider the following example, where the @alt2@ operator pulls two elements from its first input stream, then two from the second, before pushing all four to its output stream.
\begin{code}
  alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
  alternates sInA sInB sInC
   = let  s1   = alt2 sInA sInB
          s2   = alt2 sInB sInC
          sOut = zip s1 s2
     in   sOut
\end{code}

Note that the middle stream @sInB@ is shared, and the result streams from both @alt2@ operators are zipped into tuples. Given the inputs @sInA@ = @[a1,a2]@, @sInB@ = @[b1,b2]@ and @sInC@ = @[c1,c2]@ the output of @zip@ will be @[(a1,b1),(a2,b2),(b1,c1),(b2,c2)]@, assuming @a1,a2,b1,b2@ and so on are values of type @Nat@.

Now, note that the first @alt2@ process pushes values to its output stream @s1@ two at a time, and the second @alt2@ process also pushes values to its own output stream @s2@ two at a time. However, the downstream @zip@ process needs to pull one value from @s1@ then one from @s2@, then another from @s1@, then another from @s2@, alternating between the @s1@ and @s2@ streams. This will work, provided we can arrange for the two \emph{separate} @alt2@ processes to push to their separate output streams alternatively. They can still push two values at a time to their own outputs, but the downstream @zip@ process needs receive one from each process alternately. Here is a table of intermediate values to help make the explanation clearer:

\begin{code}
    sInA = [a1, a2, a3, a4, a5 ...]
    sInB = [b1, b2, b3, b4, b5 ...]
    sInC = [c1, c2, c3, c4, c5 ...]

    s1   = alt2 sInA sInB 
         = [a1, a2, b1, b2, a3, a4, b3, b4 ...]

    s2   = alt2 sInB sInC
         = [b1, b2, c1, c2, b3, b4, c3, c4 ...]

    sOut = zip s1 s2
         = [(a1,b1), (a2,b2), (b1,c1), (b2,c2) ...]
\end{code}

Considering the last line in the above table, note that @zip@ needs to output a tuple of @a1@ and @b1@ together, then @a2@ and @b2@ together, and so on. The implementation of the @zip@ process will attempt to pull the first value @a1@ from stream @s1@, blocking until it gets it, then pull the next value @b1@ from stream @s2@, blocking until it gets it. While @zip@ is blocked waiting for @b1@, the first @alt2@ process cannot yet push @a2@. The execution order of the overall process group is constrained by communication patterns of processes in that group.

As we cannot encode all possible orderings into the definition of the processes themselves, we have defined the execution rules to admit many possible orderings. In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime. In contrast, the role of our fusion transform is to construct one of these working orders statically. In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the process group was executed dynamically. In doing so, we also eliminate the need to pass messages between processes --- once they are fused we can just copy values between heap locations.

@@

% Rule (FeedPush) allows the process network to emit a push message. As with (FeedInternal), it first feeds its input accumulator, then allows the resulting network to emit a push message. The pushed value is collected in the accumulator list for that stream.

% Rule (FeedExternal) allows inputs to be injected into the process network.
% For any channel $c$ which is not an output of one of the processes, we take the last value off its list. The recursive feed is evaluated with the last value removed from the accumulators. The last value is then injected into the network, and added back to the result accumulators.

% Rule (FeedStart) applies when all input values have been injected and there are no input values left. In this case, the output values are the same as the input values.

% Evaluation: feeding evaluates a process network on a list of input values and collects the outputs. Feeding alternates between injecting input values and shaking the processes to collect the output messages.

% Note that the result stream and network are not canonical, as an infinite @push@ loop has an infinite number of evaluations.
% The feed form does not ensure that the processes themselves have finished evaluating, only that all input values have been injected.

% -- cuts ---------------------------------------------------------------------
% BL: Discuss this during def of evaluation.

% These input states are used for evaluation to ensure that communication between processes does not require unbounded buffers.

% Claim "functional dataflow" or some such. Each process only updates values in its local heap, the only intra-process communication is via streams. This is unlike StreamIt which can send out-of-band messages between its processes.

% The output streams are in some sense ``owned'' by the process that produces them: while a stream may be consumed by any number of processes, each stream can only appear as the output for one process. This ensures a sort of determinism in the scheduling of multiple processes; if different processes could push to the same stream, the order of values would depend on the scheduled order. A process may, however, produce multiple output streams.

% Each process has its own private heap, therefore the only communication between processes occurs by streams.

% The instructions (@instrs@) are a mapping from label to instruction, and label points to the current instruction. Instructions can pull from a channel, drop an already pulled value, push a value, perform an if/case analysis on a boolean, or perform an internal jump.

% After values have been pulled, they must be disposed of with @drop@: this empties the value from the buffer and allows the producer to push to the channel.

% A process network is a set of multiple processes that can be evaluated concurrently. Any inputs that are not produced as outputs of processes are assumed to be external inputs --- their values will be provided by the environment. Processes form the essence of stream computation, and a single process can be given a straightforward sequential semantics by mapping to an imperative language. By fusing multiple processes into a single one, we are effectively giving a sequential interpretation for concurrent processes.

>>>>

A $\Proc$ is a record with five fields: the @ins@ field specifies the input channels; the @outs@ field the output channels; the @heap@ field the process-local heap; the @label@ field the label of the instruction currently being executed, and the @instrs@ a map of labels to instructions. We use the same record when specifying both the definition of a particular process, as well as when giving the evaluation semantics. When specifying a process the @label@ field gives the entry-point to the process code, though during evaluation it is the label of the instruction currently being executed. Likewise, when specifying a process we usually only list channel names in the @ins@ field, though during evaluation they are also paired with their current $\InputState$. If an $\InputState$ is not specified we assume it is `none'.
A network is a set of processes, able to communicate to each other.

@@

Instructions include a $\Next$ field containing the label of the next instruction to execute, as well as a list of $\Var \times \Exp$ bindings used to update the heap.
The list of update bindings is attached directly to instructions to make the fusion algorithm easier to specify, in contrast to a presentation with a separate @update@ instruction.

@@

\begin{enumerate}
\item \emph{Injection} of a single value from a stream into a process, or a network. Each individual process only needs to accept the value when it is ready for it, and injection into a network succeeds only when they \emph{all} accept it.

\item \emph{Advancing} a single process from one state to another. Advancing a network suceeds when \emph{any} of the processes in the network can advance.

@@

Execution of a network is non-deterministic. At any moment several processes may be able to take a step, while others are blocked. As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking. This enables the overall sequence of values on each output channel to be deterministic. Unlike Kahn processes, pushing to a channel can also block. Each consumer has a single element buffer, and pushing only succeeds when that buffer is empty.

@@

Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions is not. When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the network to advance without requiring unbounded buffering.

@@

% -----------------------------------------------------------------------------
\subsubsection{Injection}
Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes. The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''. The @injects@ form is similar, operating on a network.

@@

Rule (InjectMany) attempts to inject a single value into a network. We use the single process judgment form to inject the value into all processes, which must succeed for all of them.
% Once a value has been injected into all consuming processes that require it, the producing process no longer needs to retain it.

@@

Rule (Pull) takes the @pending@ value $v$ from the channel state and produces a heap update to copy this value into the variable $x$ in the @pull@ instruction. We use the syntax $us,x=v$ to mean that the list of updates $us$ is extended with the new binding $x=v$. In the result channel states, the state of the channel $c$ that was pulled from is set to @have@, to indicate the value has been copied into the local variable.

@@

The statement $\ProcShake{p}{a}{p'}$ reads ``process $p$ advances to new process $p'$, yielding action $a$''. Rule (Advance) advances a single process. We look up the current instruction for the process @label@ and pass it, along with the channel states and heap, to the above single instruction judgment. The update expressions $us$ from the single instruction judgment are reduced to values before updating the heap. We use $(us \lhd bs)$ to replace bindings in $us$ with new ones from $bs$. As the update expressions are pure, the evaluation can be done in any order.

% -----------------------------------------------------------------------------
\subsubsection{Feeding}
Fig.~\ref{fig:Process:Eval:Feed} gives the rules for collecting output actions and feeding the contained values to other processes. The first set of rules concerns feeding values to other processes within the same network, while the second exchanges input and output values with the environment the network is running in.

The statement $\ProcShake{ps}{a}{ps'}$ reads ``the network $ps$ advances to the network $ps'$ yielding output action $a$''.

Rule (ProcessInternal) allows an arbitrary process in the network to advance to a new state at any time, provided it does not yield an output action. This allows processes to perform internal computation.
% , without needing to synchronize with the rest of the network.

Rule (ProcessPush) allows an arbitrary process in the network to advance to a new state while yielding an output action $(@push@~c~v)$. For this to succeed it must be possible to inject the output value $v$ into all processes that have channel $c$ as one of their inputs.
% As all consuming processes must accept the output value at the time it is created, there is no need to buffer it further in the producing process. When any process in the network produces an output action we take that as the action of the whole network.

The statement $\ProcsFeed{cvs}{ps}{cvs'}{ps'}$ reads ``with channel values $cvs$, network $ps$ takes a step and produces new channel values $cvs'$ and network $ps'$''. The channel values $cvs$ map channel names to a list of values. For input channels of the overall network, we initialize the map to contain a list of input values for each channel. For output channels of the overall network, values pushed to those channels are also collected in the same channel map. In a concrete implementation the input and output values would be transported over some IO device, but for the semantics we describe the abstract behavior only.

Rule (FeedInternal) allows the network to perform local computation in the context of the channel values.

Rule (FeedPush) collects an output action (@push@ $c$ $v$) produced by a network and appends the value $v$ to the list corresponding to the output channel $c$.

Rule (FeedExternal) injects values from the external environment. This rule also has the side condition that values cannot be injected from the environment into output channels that are already owned by some process. This constraint is required for correctness proofs, but can be ensured by construction in a concrete implementation.
% The topology of the dataflow network does not change at runtime, so it only needs to be checked once, before execution.

% -----------------------------------------------------------------------------
\subsection{Non-deterministic Execution Order}
\label{s:EvaluationOrder}

The execution rules of Fig. \ref{fig:Process:Eval:Feed} are non-deterministic in several ways. Rule (ProcessInternal) allows any process to perform internal computation at any time, without synchronizing with other processes in the network; (ProcessPush) allows any process to perform a push action at any time, provided all other processes in the network are ready to accept the pushed value; (FeedExternal) also allows new values to be injected from the environment, provided all processes that use the channel are ready to accept the value.

In our system, allowing the execution order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering.
In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime. In contrast, the role of our fusion transform is to construct one of these working orders statically. In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the network were executed dynamically. In doing so, we also eliminate the need to pass messages between processes --- once they are fused we can just copy values between heap locations.

% In our system, allowing the execution order of processes to be non-deterministic is critical, as it provides freedom to search for a valid ordering that does not require excessive buffering. Consider the following example, where the @alt2@ operator pulls two elements from its first input stream, then two from the second, before pushing all four to its output stream.
% \begin{code}
%   alternates : S Nat -> S Nat -> S Nat -> S (Nat, Nat)
%   alternates sInA sInB sInC
%    = let  s1   = alt2 sInA sInB
%           s2   = alt2 sInB sInC
%           sOut = zip s1 s2
%      in   sOut
% \end{code}
% 
% Note that the middle stream @sInB@ is shared, and the result streams from both @alt2@ operators are zipped into tuples. Given the inputs @sInA@ = @[a1,a2]@, @sInB@ = @[b1,b2]@ and @sInC@ = @[c1,c2]@ the output of @zip@ will be @[(a1,b1),(a2,b2),(b1,c1),(b2,c2)]@, assuming @a1,a2,b1,b2@ and so on are values of type @Nat@.
% 
% Now, note that the first @alt2@ process pushes values to its output stream @s1@ two at a time, and the second @alt2@ process also pushes values to its own output stream @s2@ two at a time. However, the downstream @zip@ process needs to pull one value from @s1@ then one from @s2@, then another from @s1@, then another from @s2@, alternating between the @s1@ and @s2@ streams. This will work, provided we can arrange for the two \emph{separate} @alt2@ processes to push to their separate output streams alternatively. They can still push two values at a time to their own outputs, but the downstream @zip@ process needs receive one from each process alternately. Here is a table of intermediate values to help make the explanation clearer:
% 
% \begin{code}
%     sInA = [a1, a2, a3, a4, a5 ...]
%     sInB = [b1, b2, b3, b4, b5 ...]
%     sInC = [c1, c2, c3, c4, c5 ...]
% 
%     s1   = alt2 sInA sInB 
%          = [a1, a2, b1, b2, a3, a4, b3, b4 ...]
% 
%     s2   = alt2 sInB sInC
%          = [b1, b2, c1, c2, b3, b4, c3, c4 ...]
% 
%     sOut = zip s1 s2
%          = [(a1,b1), (a2,b2), (b1,c1), (b2,c2) ...]
% \end{code}
% 
% Considering the last line in the above table, note that @zip@ needs to output a tuple of @a1@ and @b1@ together, then @a2@ and @b2@ together, and so on. The implementation of the @zip@ process will attempt to pull the first value @a1@ from stream @s1@, blocking until it gets it, then pull the next value @b1@ from stream @s2@, blocking until it gets it. While @zip@ is blocked waiting for @b1@, the first @alt2@ process cannot yet push @a2@. The execution order of the overall network is constrained by communication patterns of processes in that network.

% As we cannot encode all possible orderings into the definition of the processes themselves, we have defined the execution rules to admit many possible orderings. In a direct implementation of concurrent processes using message passing and operating system threads, an actual, working, execution order would be discovered dynamically at runtime. In contrast, the role of our fusion transform is to construct one of these working orders statically. In the fused result process, the instructions will be scheduled so that they run in one of the orders that would have arisen if the network was executed dynamically. In doing so, we also eliminate the need to pass messages between processes --- once they are fused we can just copy values between heap locations.

@@

### 2017-05-04 2017mergingmerges/section/04-Fusion.tex
serious cutting

The core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process group we fuse successive pairs of processes until only one remains.

Fig.~\ref{fig:Fusion:Types} defines some auxiliary grammar used during fusion. We extend the exiting $Label$ grammar with an additional alternative, $Label_S \times Label_S$ which we use for the labels in a fused result process. Each $Label_S$ consists of a $Label$ from one of the  source processes, paired with a map from $Channel$ to the statically known part of that channel's current $\InputState$. The definition of $Label$ is now recursive. While fusing a whole process group, as we fuse pairs of individual processes the labels in the result collect more and more information. Each label of the final, completely fused process encodes the joint state that all the original source processes would be in at that point.

% These new labels $Label_S$ consist of a pair of source labels, as well as the static part of the $\InputState$ of each input channel.

% If the static $\InputState_S$ is $@pending@_S$, there is a value waiting to be pulled, do not know the actual value.

We also extend the existing $Variable$ grammar with a (@chan@ c) form which represents the stream buffer variable associated with channel @c@. We only need one stream buffer variable for each channel, and naming them like this saves us from inventing fresh variable names in the definition of the fusion rules (we used a fresh name @b1@ back in \S\ref{s:Fusion:FusingPulls} to avoid introducing a new mechanism at that point in the discussion)

Still in Fig.~\ref{fig:Fusion:Types}, $\ChanType_2$ classifies how channels are used, and possibly shared, between two processes. Type @in2@ indicates that the two processes @pull@ from the same channel, and these actions must then be coordinated. Type @in1@ indicates that only a single process pulls from the channel. Type @in1out1@ indicates that one process pushes to the channel and the other pulls from the same. Type @out1@ indicates that the channel is pushed to by a single process. Each output channel is uniquely owned and cannot be pushed to by more than one process.

@@

\smallskip
% -------------------------------------
Fig.~\ref{fig:Fusion:Def:Top} defines function \ti{fusePair} that fuses a pair of processes together, constructing a result process that does the job of both. We start with a joint label $l_0$ formed from the initial labels of the two source processes. We then use \ti{tryStepPair} to statically choose which of the two processes to advance, and hence which instruction to execute next. The possible destination labels of that instruction (computed with $outlabels$ from Fig.~\ref{fig:Fusion:Utils}) define new joint labels and reachable states. As we discover reachable states we add them to a map $bs$ of joint label to corresponding the instruction, and repeat the process to a fixpoint where no more states can be discovered.

@@

Clauses (PreferJump1) and (PreferJump2) are heuristics which prioritize processes that can perform a @jump@. This heuristic helps to collect @jump@ instructions together in the result process, which are then easier for post-fusion optimization (\S\ref{s:Optimisation}) to handle. As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ (from Fig.~\ref{fig:Fusion:Utils}) to re-swap the labels in the result.

Clauses (DeferPull1) and (DeferPull2) are similar heuristics, which defer @pull@ instructions if possible. We do this because @pull@ instructions may block, while other instructions are more likely to produce immediate results.

@@

Clause (LocalCase) is similar, except that there are two destination labels.

@@

Clause (SharedPush) applies when the left process wants to push to a shared channel. Pushing to a shared channel requires the downstream process to be ready to accept the value at the same time. We encode this constraint by requiring the static input state of the downstream channel to be $@none@_S$. When this is satisfied, the result instruction stores the pushed value in the stream buffer variable $@chan@~c$'' and sets the static input state to ``$@pending@_S$'', which indicates that the new value is now available.

@@

Clause (SharedPull) applies when the left process wants to pull from a shared channel that the other process either pulls from or pushes to. We know that there is already a value in the stream buffer variable, because the state for that channel is $@pending@_S$. The result instruction copies the value from the stream buffer variable into a variable specific to the left source process, and the corresponding $@have@_S$ channel state in the result label records that it has done so.

@@

Clauses (SharedDropOne) and (SharedDropBoth) apply when the left process wants to drop a value it gained from a channel which is shared by the other process. In (SharedDropOne) the channel states reveal that the other process is still using the value. In this case the result instruction is a @jump@ that only updates the static information that the left source process no longer requires this value. In (SharedDropBoth) then channel states reveal that the other process no longer needs the value either. In this case the result instruction is a real @drop@, because we are sure that neither process requires the value any longer.

@@

\smallskip
Fig.~\ref{fig:Fusion:Utils} contains definitions of some utility functions which we have already mentioned.
Function \ti{channels} computes the $\ChanType_2$ map for a pair of processes.
Function \ti{outlabels} gets the set of output labels for an instruction, which is used when computing the fixpoint of reachable states.
Function \ti{swaplabels} flips the order of the compound labels in an instruction.

>>>>

The core fusion algorithm constructs a static execution schedule for a single pair of processes. To fuse a whole process network we fuse successive pairs of processes until only one remains.

Fig.~\ref{fig:Fusion:Types} defines some auxiliary grammar used during fusion. We extend the existing $Label$ grammar with an additional alternative, $Label_F \times Label_F$ for the labels in a fused result process. Each $Label_F$ consists of a $Label$ from one of the source processes, paired with a map from $Channel$ to the statically known part of that channel's current $\InputState$. The definition of $Label$ is now recursive. While fusing a whole network, as we fuse pairs of individual processes the labels in the result collect more and more information. Each label of the final, completely fused process encodes the joint state that all the original source processes would be in at that point.

% These new labels $Label_F$ consist of a pair of source labels, as well as the static part of the $\InputState$ of each input channel.

% If the static $\InputState_F$ is $@pending@_F$, there is a value waiting to be pulled, do not know the actual value.

We also extend the existing $\Var$ grammar with a (@chan@ $c$) form which represents the buffer variable associated with channel @c@. We only need one buffer variable for each channel, and naming them like this saves us from inventing fresh names in the definition of the fusion rules.
(We used a fresh name back in \S\ref{s:Fusion:FusingPulls} to avoid introducing a new mechanism at that point in the discussion.)

Still in Fig.~\ref{fig:Fusion:Types}, $\ChanType_2$ classifies how channels are used, and possibly shared, between two processes. Type @in2@ indicates that the two processes @pull@ from the same channel, and these actions must then be coordinated. Type @in1@ indicates that only a single process pulls from the channel. Type @in1out1@ indicates that one process pushes to the channel and the other pulls. Type @out1@ indicates that the channel is pushed to by a single process. Each output channel is uniquely owned and cannot be pushed to by more than one process.

@@

\smallskip
% -------------------------------------
Fig.~\ref{fig:Fusion:Def:Top} defines function \ti{fusePair} that fuses a pair of processes together, constructing a result process that does the job of both. We start with a joint label $l_0$ formed from the initial labels of the two source processes. We then use \ti{tryStepPair} to statically choose which of the two processes to advance, and hence which instruction to execute next. The possible destination labels of that instruction (computed with $outlabels$ from Fig.~\ref{fig:Fusion:Utils}) define new joint labels and reachable states. As we discover reachable states we add them to a map $bs$ of joint label to corresponding the instruction, and repeat the process to a fixpoint where no new states can be discovered.

@@

Clauses (PreferJump1) and (PreferJump2) are heuristics which prioritize processes that can perform a @jump@. This helps collect @jump@ instructions together, which are then easier for post-fusion optimization (\S\ref{s:Optimisation}) to handle. As clause (PreferJump2) calls $\ti{tryStep}$ with the instructions in swapped order, if \ti{tryStep} succeeds we use $\ti{swaplabels}$ (from Fig.~\ref{fig:Fusion:Utils}) to re-swap the labels in the result.

Similarly, clauses (DeferPull1) and (DeferPull2) defer @pull@ instructions. We do this because @pull@ instructions may block, while other instructions are more likely to produce immediate results.

@@

Clause (LocalCase) is similar, except there are two $\Next$ labels.

@@

Clause (SharedPush) applies when the left process wants to push to a shared channel. Pushing to a shared channel requires the downstream process to be ready to accept the value at the same time. We encode this constraint by requiring the static input state of the downstream channel to be $@none@_F$. When this is satisfied, the result instruction stores the pushed value in the stream buffer variable $(@chan@~c)$ and sets the static input state to $@pending@_F$, which indicates that the new value is now available.

@@

Clause (SharedPull) applies when the left process wants to pull from a shared channel that the other process either pulls from or pushes to. We know that there is already a value in the stream buffer variable, because the state for that channel is $@pending@_F$. The result instruction copies the value from the stream buffer variable into a variable specific to the left source process, and the corresponding $@have@_F$ channel state in the result label records that it has done so.

@@

Clauses (SharedDropOne) and (SharedDropBoth) apply when the left process wants to drop from a channel shared by both processes. In (SharedDropOne) the channel states reveal that the other process is still using the value. In this case the result is a @jump@ updating the channel state to note that the left process has dropped. In (SharedDropBoth) the channel states reveal that the other process no longer needs the value. In this case the result is a real @drop@, because we are sure that neither process requires the value any longer.

@@

\smallskip
Fig.~\ref{fig:Fusion:Utils} contains definitions of utility functions which we have already mentioned.
Function \ti{channels} computes the $\ChanType_2$ map for a pair of processes.
Function \ti{outlabels} gets the set of output labels for an instruction, which is used when computing the fixpoint of reachable states.
Function \ti{swaplabels} flips the order of the compound labels in an instruction.
